A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-08 latex/main.tex
Updates from Overleaf

\section*{INTRODUCTION}
% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
% in terms of the basic goal, assumptions about samples and how recombination
% can be treated.
Classical phylogenetics is largely concerned with inferring the evolutionary history 
of species~\cite{Felsenstein2004}, with small numbers of widely diverged
samples. In this context, recombination is a nuisance which must be 
carefully accounted for~\cite{Rannala2025recombination},
affecting the inference of species trees~\cite{Posada2002effect}
and biasing downstream
analyses~\cite{Schierup2000consequences,Hedge2014bacterial}.
Modern pathogen datasets are not sparsely sampled from different species,
but consist of tens to hundreds of thousands of samples from 
a single species~\cite{
Shu2017,% GISAID ~270k FLU + others
Jolley2018open,% PubMLST, 34K genomes for e.g. Staph Auerus
Apetrei2021hiv,% HIV 1,026,217 sequences
Cryptic2022data,% TB 12k
Dyer2025enterobase% 1 million, various things like E-coli
}.
In contrast to the sophisticated model-based approaches
used to perform species-level tree
inference~\cite{Suchard2018bayesian,Bouckaert2019beast,Minh2020iq},
comprehensive phylogenies of these datasets are typically built with
simpler distance methods [citation?]
or approximations~\cite{Price2009fasttree},
primarily due to computational scaling limitations,
but also because of the complexity of modeling such datasets.
While several methods have been proposed to incorporate
recombination explicitly into model-based phylogenetic
inference~\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Muller2022bayesian},
scalability
is a major issue and practical application has been limited.
Software support in explicitly \emph{using} recombination-aware
inference is also extremely limited
[TODO add some concrete details].

@@

[TODO we should mention UShER here: they state multiple times in the GitHub issues 
that they don't handle them at all, so I think we can just say this in some nice way.]
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
but because they can be difficult to distinguish from bioinformatics errors
and their multi-base nature presents challenges
to the Markovian structure of the LS HMM,
we masked out all gap characters as missing data during primary inference.
To incorporate the most important deletions into the ARG,
we remapped data using post hoc parsimony at 75 sites involved in deletions
with a frequency of $\geq 1\%$ reported by Li et al.\cite{Li2023indels}
during post-processing (STAR Methods).
Although sites are treated independently in this approach,
runs of mutations to the gap character at adjacent sites
correctly identified many known
high-frequency deletions (Table~\ref{tab:major_dels}) and
recurrent deletions (Table~\ref{tab:rec_dels}).
For example, the three characteristic deletions of Alpha \cite{RambautAlpha2020}
(11288+9, 21765+6, 21991+3,
where the first and second numbers are
the start position and length, respectively)
are correctly mapped above the Alpha origin in the ARG
(Figure~\ref{fig:subgraph_alpha}).
The ARG captures known recurrent deletions
in the N-terminal domain of the S1 unit of Spike
(which is a deletion hotspot \cite{McCarthy2021})
for example,
the 21765+6 deletion ($\Delta$H69/V70),
which causes S-Gene Target Failure \cite{Walker2021} and
is associated with increased efficiency in cell entry \cite{Meng2021}, and
the 21991+3 deletion
which is associated with antibody escape \cite{McCarthy2021}.
See Document\ref{sec:supplement_dels} for further analysis.

@@

\subsection*{Comparison with RIPPLES}
RIPPLES~\cite{Turakhia2022} is a method for detecting 
recombination events in a tree inferred by UShER. Recombination is 
detected by first identifying plausible branches carrying multiple
mutations (at least 3, by default). For each candidate branch,
RIPPLES then systematically splits the sequence into 2 or 3
segments, and finds parsimonious placements for each segment independently. 
A recombination event at that node is then flagged if the overall parsimony of these multiple 
placements sufficiently improves over the original single-parent
origin in the tree (the ``parsimony improvement'' parameter, $p=3$, by default).
RIVET~\cite{Smith2023tracking} is a tool for quality control
and visualisation of these recombination events.

@@

Table~\ref{tab:ripples_x_lineages} shows a summary of all 
recombination events associated with Pango X lineages detected by RIPPLES,
and their relationship with the sc2ts ARG. 
Of the 16 Class I events 
(Pango X lineages unequivocally associated with a recombination node in
the sc2ts ARG) 
in Table~\ref{tab:pango_x_lineages}, 8
have corresponding recombination events (4, at $p=4$). Of the 5 Type II
events (recombination nodes with more complex assocation with X lineages),
two have close matches. XBM matches exactly (including 2 descending BF.3 samples),
and XE/XH is similar (but without including the 2 XH sample in the clade).
UShER and sc2ts agree on creating two separate XAM clades, but while sc2ts
makes these both descendants of the Xx event, RIPPLEs creates two independent
recombinations. Similarly, XAE and have exactly matching clades, but RIPPLEs
infers to independent recombination events while sc2ts groups them both
under the Xx event. XB is notable as a likely sc2ts false negative that 
identified by RIPPLES at both $p=3$ and $p=4$. The remaining events 
associated with X lineages output by RIPPLES cannot be easily mapped 
to the sc2ts ARG as one or more descendants is distantly placed.
XAJ is an outlier for sc2ts, being the Pango X lineage identified as a 
non-recombinant with the largest number of mutations 
(Table~\ref{tab:pango_x_lineages}), and a likely false negative.
The remaining 4 events are mixtures of multiple lineages that are 
difficult to reconcile with either the other events in the table 
or the sc2ts ARG.

@@

Sequencing errors in genome assemblies \cite{Turakhia2020},
read misalignments \cite{Hunt2024}, and
multi-nucleotide substitutions \cite{DeMaio2020issues}
can cause short regions of a sample genome to match more closely
to an alternative genome rather than the true parent.
Such genomic regions can be mistaken for evidence of recombination,
because the LS HMM counts each base separately.
We developed a quality control (QC) measure
which counts the number of ``loci'' (clusters of closely spaced sites)
supporting the left and right parents of a recombinant (STAR Methods),
and effectively identifies a range of issues.
Figure~\ref{fig:recombinant_qc} plots
the net number of supporting loci for the left and right parents
for all 855 recombination events.
The 501 recombinants having fewer than 4 net loci supporting each parent
are highly enriched for samples sequenced using the AmpliSeq V1 protocol
% https://github.com/jeromekelleher/sc2ts-paper/issues/808
(70\% of the recombinants despite making up 2\% of the Viridian dataset), and
71 (14\%) are associated with 6-base deletion in the Delta lineage,
known to be affected by amplicon dropout\cite{Sanderson2021}.
The 354 remaining recombination events that pass this QC filter
contain all of the recombinants associated with Pango X lineages
(Table~\ref{tab:pango_x_lineages}) and
the Jackson et al. recombinants (Document\ref{sec:jackson_recombs}).

@@

\subsubsection*{Sample filtering and preprocessing}
The collection date for a sequence is an important piece of information
for sc2ts. We used the Viridian
consensus date (``Date\_tree``), based on information from COG-UK, GISAID, and
ENA/SRA as our source. Samples that have missing or incomplete dates (e.g., 2020-01)
were omitted, as were
the $\sim$40,000 samples which had a collection date of December 31, 2020.
As the sample count for this date is a major outlier for this period of
the pandemic, it seems likely that this 
batch is highly enriched for incorrect collection dates.
Manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
confirmed that some of the samples were not collected in 2020.

The bioinformatics community has curated a set of problematic sites
for SARS-CoV-2 that are enriched for systematic sequencing errors or 
high levels of homoplasy~\cite{DeMaio2020issues}. 
Although these sites are recommended to be masked for UShER
inference~\cite{Turakhia2020},
the list has not been updated to take into account the systematic improvements 
made by the Viridian dataset, and we therefore assembled a custom set of sites to mask. 
To do so, we first built a preliminary ARG from the samples collected up to June 30, 2021,
without any site masking.
Then, we identified the top 100 sites in terms of mutation counts in the ARG.
Only 11 of these sites 
(635, 8835, 11074, 11083, 15521, 16887, 21304, 21305, 21575, 21987 and 28253)
were present in the problematic sites file
(\url{https://raw.githubusercontent.com/W-L/ProblematicSites\_SARS-CoV2/master/problematic\_sites\_sarsCov2.vcf}),
and most sites in this list were typical in terms of the numbers of mutations.
We therefore masked our 100 most-mutagenic sites to remove sequencing errors and improve
matching performance in the LS HMM.

Alignments read from the VCZ Zarr file were first preprocessed to mask any 
non nucleotide characters (A,C,G or T) as missing data. 
Then, we filtered 
any samples that had missing data at more than 500 sites, excluding
the 100 sites identified as problematic (this was 
set to 10,000 sites for the initial period up to 2020-03-01 where sampling
density is low).

@@

The LS model \cite{LiStephens2003} is an approximation of the coalescent with recombination
that captures many of the key features of the joint processes of mutation and recombination.
It is widely used in a range of applications in population genetics, including ARG 
% Does Threads use LS?
inference \cite{Kelleher2019,Gunnarsson2024}.
The LS model is an HMM that models a focal genome as a sequence of nucleotides (i.e. the observed states),
which are probabilistically emitted as an imperfect mosaic of a set of genomes in a reference panel (i.e., the hidden states).
The process of switching between the genomes in the reference panel is governed by a transition matrix,
and mismatches between a focal genome and a genome of the reference panel (from which it is ``copied'') are permitted as emissions.

>>>>

\section*{INTRODUCTION}
% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
% in terms of the basic goal, assumptions about samples and how recombination
% can be treated.
Classical phylogenetics is largely concerned with inferring the evolutionary history 
of species~\cite{Felsenstein2004}, with small numbers of widely diverged
samples. In this context, recombination is a nuisance which must be 
carefully accounted for~\cite{Rannala2025recombination},
affecting the inference of species trees~\cite{Posada2002effect}
and biasing downstream
analyses~\cite{Schierup2000consequences,Hedge2014bacterial}.
Modern pathogen datasets are not sparsely sampled from different species,
but consist of tens to hundreds of thousands of samples from 
a single species~\cite{
Shu2017,% GISAID ~270k FLU + others
Jolley2018open,% PubMLST, 34K genomes for e.g. Staph Auerus
Apetrei2021hiv,% HIV 1,026,217 sequences
Cryptic2022data,% TB 12k
Dyer2025enterobase% 1 million, various things like E-coli
}.
In contrast to the sophisticated model-based approaches
used to perform species-level tree
inference~\cite{Suchard2018bayesian,Bouckaert2019beast,Minh2020iq},
comprehensive phylogenies of these datasets are typically built with
simpler distance methods [citation?]
or approximations~\cite{Price2009fasttree},
primarily due to computational scaling limitations,
but also because of the complexity of modeling such datasets.
While several methods have been proposed to incorporate
recombination explicitly into model-based phylogenetic
inference~\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Muller2022bayesian},
scalability
is a major issue and practical application has been limited.
Software support in explicitly \emph{using}
recombination-aware inference is limited to
visualization of small-scale phylogenetic
networks~\cite{Huson2012,Vaughan2017,Solís-Lemus2017} and
using tanglegrams to visualize tree
changes (e.g.~\cite{Rasmussen2023}),
typically using the Extended Newick format for
interchange~\cite{Cardona2008}.

@@

[TODO we should mention UShER here: they state multiple times in the GitHub issues 
that they don't handle them at all, so I think we can just say this in some nice way.]
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
but because they can be difficult to distinguish from bioinformatics errors
and their multi-base nature presents challenges
to the Markovian structure of the LS HMM,
we masked out all gap characters as missing data during primary inference.
To incorporate the most important deletions into the ARG,
we remapped data using post hoc parsimony at 75 sites involved in deletions
with a frequency of $\geq 1\%$ reported by Li et al.\cite{Li2023indels}
during post-processing (STAR Methods).
Although sites are treated independently in this approach,
runs of mutations to the gap character at adjacent sites
correctly identified many known
high-frequency deletions (Table~\ref{tab:major_dels}) and
recurrent deletions (Table~\ref{tab:rec_dels}).
For example, the three characteristic deletions of Alpha \cite{RambautAlpha2020}
(11288+9, 21765+6, 21991+3,
where the first and second numbers are
the start position and length, respectively)
are correctly mapped above the Alpha origin in the ARG
(Figure~\ref{fig:subgraph_alpha}).
The ARG captures known recurrent deletions
in the N-terminal domain of the S1 unit of Spike
(which is a deletion hotspot \cite{McCarthy2021})
for example,
the 21765+6 deletion ($\Delta$H69/V70),
which causes S-Gene Target Failure \cite{Walker2021} and
is associated with increased efficiency in cell entry \cite{Meng2021}, and
the 21991+3 deletion
which is associated with antibody escape \cite{McCarthy2021}.
See Document~\ref{sec:supplement_dels} for further analysis.

@@

\subsection*{Comparison with RIPPLES}
RIPPLES~\cite{Turakhia2022} is a method for detecting 
recombination events using a tree inferred by UShER. Recombination is 
detected by first identifying plausible branches carrying multiple
mutations (at least 3, by default). For each candidate branch,
RIPPLES then systematically splits the sequence into 2 or 3
segments, and finds parsimonious placements for each segment independently. 
A recombination event at that node is then flagged if the overall parsimony of these multiple 
placements sufficiently improves over the original single-parent
origin in the tree (the ``parsimony improvement'' parameter, $p=3$, by default).
RIVET~\cite{Smith2023tracking} is a tool for quality control
and visualisation of these recombination events.

@@

Table~\ref{tab:ripples_x_lineages} shows a summary of all 
recombination events associated with Pango X lineages detected by RIPPLES,
and their relationship with the sc2ts ARG. 
Of the 16 Class I events 
(Pango X lineages unequivocally associated with a recombination node in
the sc2ts ARG) 
in Table~\ref{tab:pango_x_lineages}, 8
have corresponding recombination events (4, at $p=4$). Of the 5 Type II
events (recombination nodes with more complex assocation with X lineages),
two have close matches. XBM matches exactly (including 2 descending BF.3 samples),
and XE/XH is similar (but without including the 2 XH samples in the clade).
UShER and sc2ts agree on creating two separate XAM clades, but while sc2ts
makes these both descendants of the Xx event, RIPPLES creates two independent
recombinations. Similarly, XAE and ??? have exactly matching clades, but RIPPLES
infers two independent recombination events while sc2ts groups them both
under the Xx event. XB is notable as a likely sc2ts false negative that is 
identified by RIPPLES at both $p=3$ and $p=4$. The remaining events 
associated with X lineages output by RIPPLES cannot be easily mapped 
to the sc2ts ARG as one or more descendants is distantly placed.
XAJ is an outlier for sc2ts, being the Pango X lineage identified as a 
non-recombinant with the largest number of mutations 
(Table~\ref{tab:pango_x_lineages}), and a likely false negative.
The remaining 4 events are mixtures of multiple lineages that are 
difficult to reconcile with either the other events in the table 
or the sc2ts ARG.

@@

Sequencing errors in genome assemblies \cite{Turakhia2020},
read misalignments \cite{Hunt2024}, and
multi-nucleotide substitutions \cite{DeMaio2020issues}
can cause short regions of a sample genome to match more closely
to an alternative genome rather than the true parent.
Such genomic regions can be mistaken for evidence of recombination,
because the LS HMM counts each base separately.
We developed a quality control (QC) measure
which counts the number of ``loci'' (clusters of closely spaced sites)
supporting the left and right parents of a recombinant (STAR Methods),
and effectively identifies a range of issues.
Figure~\ref{fig:recombinant_qc} shows
the net number of supporting loci for the left and right parents
for all 855 recombination events.
The 501 recombinants having fewer than 4 net loci supporting each parent
are highly enriched for samples sequenced using the AmpliSeq V1 protocol
% https://github.com/jeromekelleher/sc2ts-paper/issues/808
(70\% of the recombinants despite making up 2\% of the Viridian dataset), and
71 (14\%) are associated with 6-base deletion in the Delta lineage,
known to be affected by amplicon dropout\cite{Sanderson2021}.
The 354 remaining recombination events that pass this QC filter
contain all of the recombinants associated with Pango X lineages
(Table~\ref{tab:pango_x_lineages}) and
the Jackson et al. recombinants (Document\ref{sec:jackson_recombs}).

@@

\subsubsection*{Sample filtering and alignment preprocessing}

The collection date of a sample
is an important piece of information for sc2ts.
We used the Viridian consensus date (``Date\_tree``),
based on information from COG-UK, GISAID, and ENA/SRA as our source.
Samples which have missing or incomplete dates (e.g., 2020-01) were omitted,
as were the $\sim$40,000 samples
which had a collection date of December 31, 2020.
As the sample count for this date is a major outlier
for this period of the pandemic,
it seems likely that this batch is highly enriched
for incorrect collection dates.
Manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
confirmed that some of the samples were not collected in 2020.

The bioinformatics community has curated a set of problematic sites
for SARS-CoV-2 that are enriched for systematic sequencing errors or 
high levels of homoplasy~\cite{DeMaio2020issues}. 
Although these sites are recommended to be masked for UShER
inference~\cite{Turakhia2020},
the list has not been updated to take into account
the systematic improvements made by the Viridian dataset, and
we therefore assembled a custom set of sites to mask. 
To do so, we first built a preliminary ARG
from the samples collected up to June 30, 2021,
without any site masking.
Then, we identified the top 100 sites
in terms of mutation counts in the ARG.
Only 11 of these sites
(635, 8835, 11074, 11083, 15521, 16887,
21304, 21305, 21575, 21987 and 28253)
were present in the problematic sites file
(\url{https://github.com/W-L/ProblematicSites_SARS-CoV2}),
and most sites in this list were typical
in terms of the numbers of mutations.
We therefore masked our 100 mutation-rich sites
to remove sequencing errors and
improve matching performance in the LS HMM.

Alignments read from the VCZ Zarr file were first preprocessed
to mask any non nucleotide characters (A,C,G or T) as missing data.
% https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian_config.toml
Then, we filtered any samples that had missing data at more than 500 sites,
excluding the 100 sites identified as problematic
(this was set to 10,000 sites for the initial period up to 2020-03-01
where sampling density is low).

@@

The LS model \cite{LiStephens2003} is an approximation of
the coalescent with recombination
that captures many of the key features of
the joint processes of mutation and recombination.
It is widely used in a range of applications in population genetics,
% Does Threads use LS?
including ARG inference \cite{Kelleher2019,Gunnarsson2024}.
The LS model is an HMM that models a focal genome
as a sequence of nucleotides (i.e. the observed states),
which are probabilistically emitted
as an imperfect mosaic of a set of genomes
in a reference panel (i.e., the hidden states).
The process of switching between the genomes in the reference panel
is governed by a transition matrix,
and mismatches between a focal genome and a genome of the reference panel
(from which it is ``copied'') are permitted as emissions.
